Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-07-05 introduction.tex
lloid_paper: reworked EM sources section to be more complete and specific
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1266 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\textsc{ns}) is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  Effort from both the \GW{} and astronomy communities may make it
possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the
\GW{} community initiated a project to send alerts when potential \GW{}
transients were observed in order to trigger followup observations by \EM{}
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011},
which was an important achievement, but too late to catch any prompt \EM{}
emission.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one may have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of this
problem, a computationally inexpensive low-latency filtering algorithm for detecting
inspiral signals in \GW{} data.  We also consider the prospects for advanced \GW{}
detectors and discuss other areas of work that would be required for rapid analysis.

Compact binary coalescence (\CBC) events are thought to be a mechanism for
short gamma-ray bursts (\GRB{}s)~\citep{Lee:2005, nakar07}.  In this
scenario, prompt gamma-ray emission arises on the accretion timescale of a
central compact object formed $0.1$--$1.0$~s after the merger~\citep{Janka1999}.
Optical flashes have only been observed for a handful of long \GRB{}s
\citep{2011CRPhy..12..255A} by telescopes with extremely rapid response or, in
the case of \textsc{grb 080319b}, by pure serendipity, where several telescopes
were already observing the afterglow of another \GRB{} in the same field of
view \citep{2008Natur.455..183R}. The observed optical flashes peaked within
tens of seconds and decayed quickly.  Short \GRB{}s, on the other hand,
typically fade so quickly that it is difficult to catch even the tails of the
afterglows in any band. Rapid \GW{} transient alerts could enable the
observation of prompt optical flashes and the rise of afterglows from short
\GRB{}s.

Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended emission of $30$--$100$\,s in duration beginning
$\sim$$10$\,s after the \GRB{} and carrying comparable fluence to the initial
outburst.  Potential explanations for the emission are delayed fall-back
accretion caused by $r$-process heating \citep{Metzger2010} or the formation of
a proto-magnetar that interacts with ejecta \citep{Bucciantini2011}.  A
proto-magnetar may form from accretion-induced collapse or the merger of two
\textsc{ns}s.  An accompanying \GW{} observation would confirm the existence of
the latter channel.

@@

There were a number of sources of latency associated with \CBC{}
analysis in S6/VSR3 \citep{HugheyGWPAW2011}, listed here.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\textsc{ns}) is involved, it may become tidally disrupted near
the merger and fuel an electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  Effort from both the \GW{} and astronomy communities may make it
possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the
\GW{} community initiated a project to send alerts when potential \GW{}
transients were observed in order to trigger followup observations by \EM{}
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011},
which was an important achievement, but too late to catch any prompt \EM{}
emission.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one may have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of this
problem, a computationally inexpensive low-latency filtering algorithm for detecting
inspiral signals in \GW{} data.  We also consider the prospects for advanced \GW{}
detectors and discuss other areas of work that would be required for rapid analysis.

Compact binary coalescence (\CBC{}) is a plausible progenitor for most short
gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}. The tidally
disrupted material falls onto the newly formed, rapidly spinning compact object
and is accelerated in jets along the spin axis with a timescale of
$0.1$--$1.0$~s after the merger \citep{Janka1999}, matching the short \GRB{}
duration distribution well. Prompt \EM{} emission can arise as fast outflowing
matter collides with slower matter ejected earlier in inner shocks. The same
inner shocks, or potentially reverse shocks, can produce an accompanying
optical flash \citep{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which arise in
the external shock with the local medium and are relatively insensitive to
initial conditions. Optical flashes have been observed for a handful of long
\GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were already observing the afterglow of another \GRB{} in the same
field of view \citep{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly. For short \GRB{} parameters,
however, the reverse shock model predicts a peak flux in radio, approximately
20 minutes after the \GRB{}, but also a relatively faint optical flash
\citep{nakar07}; for a once-per-year Advanced LIGO event at $130$\,Mpc, the
radio flux will peak around $9$\,GHz with $\sim$$5$\,mJy and the optical at
$R\sim19$\,mag.  Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended X-ray emission of $30$--$100$\,s in duration
beginning $\sim$$10$\,s after the \GRB{} and carrying comparable fluence to the
initial outburst. This may be caused if the merger results in the formation of
a proto-magnetar that interacts with ejecta \citep{Bucciantini2011}. Rapid
\GW{} alerts would enable joint \EM{} and \GW{} observations to confirm the
existence of \CBC{} progenitors to short \GRB{}s and allow the early \EM{}
observation of exceptionally nearby and thus bright events.

@@

There were a number of sources of latency associated with the search for
\CBC{} signals in S6/VSR3 \citep{HugheyGWPAW2011}, listed here.
